Tools for identifying unexpectedly low microfossil count sums

### 2022-12-31 manuscript/count_check_MS.qmd
rmd -> qmd

---
title: "Tools for identifying unexpectedly low microfossil count sums"
author:
  - name: Richard J. Telford
    email: richard.telford@uib.no
    affiliation: University of Bergen
    footnote: Corresponding Author
address:
  - code: University of Bergen
    address: Department of Biological Sciences, University of Bergen and Bjerknes Centre for Climate Research, Post Box 7803, N-5020 Bergen, Norway
abstract: |
  Microfossil counts are a key data type in palaeoecology. Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so, and perhaps other aspects of the counts. This paper explores the statistical properties of ecological and palaeoecological species counts and uses the findings to develop methods to flag count or percentage data with unexpected distributions. Several assemblage datasets that fail these tests, and potentially have misreported counts or count sums, are identified.
journal: "Journal of Quaternary Science"
date: "`r format(Sys.Date(), '%d %B %Y')`"
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{pdflscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}   
output:
  bookdown::pdf_book:
    base_format: 
      rticles::elsevier_article
    number_sections: true
documentclass: article
classoption: a4paper
bibliography: extra/countMS2.bib
csl: extra/elsevier-harvard_rjt.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE, warning = TRUE, message = TRUE, error = TRUE)

# Packages loaded for convenience when knitting without running r_make
library("drake")
library("tidyverse")
library("rjt.misc")

@@

loadd(bird_singleton_summary, bird_summ)
loadd(testate_summ)
loadd(last_chance_summ)
loadd(pollen_summ, hershop_pollen_summ, wildhorse, pollen_summ1, pollen_summ2, pollen3, pollen1, pollen2)
loadd(owens_summ1, owens_summ, owens_est_summ)
loadd(diatom1_summ, diatom2_summ, chironomid1_summ, marine1_summ, chironomid2_summ)

@@

Count data analysed include the 1) North American Breeding Bird survey [@bbsAssistant2], a very large, annual monitoring programme to track the status of North American bird populations.
The survey reports all birds seen in point counts along ~40 km survey routes at the height of the breeding season. 
2) The Brown University planktic foraminiferal database [@Prell1999].
3) The MOLTEN diatom database of modern coastal diatom assemblages [@Clarke2006].
4) The North American testate amoeba training set [@AMESBURY2018].
5) Fossil pollen assemblage from the Neotoma database [@neotoma_2018] downloaded on `r readd(pollen_download_date)`.

@@

Across all the datasets, only a small minority of samples lack singletons, and even fewer have a greatest common divisor larger than one (Table \@ref(tab:multitaxa)).
Most of the samples without singletons, or with a greatest common divisor larger than one, have low richness (Fig. \@ref(fig:richness-singleton-plot)).

Although the proportion of pollen assemblages that lack singletons or have a greatest common divisor of one are low, such assemblages are not randomly distributed among the `r pollen_summ$n_datasets %>% format()` pollen datasets. 
Half of the assemblages without singletons are in just `r pollen_summ$half_no_sing` datasets.
`r Words(pollen_summ$half_gcd)` datasets contain half of the assemblages with a greatest common divisor above one and only `r pollen_summ$dataset_gcd1`% of the datasets have any such assemblages.
Some pollen datasets with unexpected properties are discussed in section \@ref(unexpected-count-data).

```{r multitaxa}
readd(count_summary) %>% 
  #manually escape % as cannot use kbl(escape = TRUE) with linebreak
  mutate(across(where(is.character), str_replace, pattern = "%", replacement = "\\\\%")) %>% 
  kbl(x = ., booktabs = TRUE, escape = FALSE,
      caption = "Properties of different data sets", 
      col.names = linebreak(str_wrap(names(.), width = 10))
    ) %>% 
  kable_styling(latex_options = c("basic", "hold_position"))
```

```{r richness-singleton-plot, fig.cap = "Relationship betwen the presence of singletons and taxonomic richness for the different datasets. Samples with a greatest common divisor greater than one are shown in dark blue. Curve is fitted with a generalised additive model", fig.height = 120/24.5, fig.width=160/24.5}
readd(richness_singleton_plot)
```

@@

To explore the effect of taxonomic richness on the prevalence of singletons, I aggregate the bird data to progressively lower taxonomic resolutions.
As richness declines, the proportion of counts without singletons increases (Table \@ref(tab:multitaxa)), reaching a moderate proportion at the order level, where most of the counts lacking singletons have fewer than five taxa.
Even when richness is low, only a small minority of counts have a greatest common divisor greater than one.

@@

To test the sensitivity of the prevalence of counts without singletons to the count sum, I resample the bird counts with 400 or more individuals to smaller count sums using a multivariate hypergeometric distribution.
At the order level, the proportion of counts without singletons decreases steeply until it reaches a minimum at when the count sum is about 200, thereafter it increases slowly (Fig. \@ref(fig:bird-recount)).
The prevalence of counts without singletons is high for small counts because it is possible that only the common taxa are counted; with larger counts, the chance of counting a single individual of a rare taxon increases.
With even larger counts the prevalence of routes without singletons increases due to saturation as singletons become doubletons and few new taxa are found.
The latter effect may largely be an artefact as in the simulation, the list of possible taxa to sample is short, whereas in the wild, there are many rare taxa which might be observed.

```{r bird-recount, fig.cap="Effect of count sum on the proportion of bird counts from the North American breeding bird survey without singleton at the order level. Results are the mean of ten trials with 95% confidence interval.", fig.width = 80/25.4, fig.height = 70/25.4}
readd(bird_recount_singleton_plot)
```

@@

The `r last_chance_summ$nsamples` chironomid head capsule assemblage dataset from Last Chance Lake [@Axford2017] also includes both the count sums and the percentage of each taxa.
The percentages are given to two decimal places: to account for rounding errors, the countSum package uses the smallest and largest values consistent with the reported percentage.
With the minimum percent method the estimated count sums are, within error, either identical to the reported count sum or twice as much (Fig. \@ref(fig:last-chance-plot)).
With the direct search method, the estimated count sums are all exactly twice the reported count sum.
The factor of two difference is because all the counts include half head capsules (but only sometimes is the rarest taxon represented by a half head capsule).
Reporting half microfossils is common for several microfossil groups, including chironomids and pollen (especially for bisaccate conifers); occasionally other fractions are reported.
Half counts make the estimated count too high by a factor of two, so do not risk incorrectly flagging count sums as being too small.

```{r last-chance-plot, fig.cap = "Estimated and reported chironomid count sums by the minumum percent method (solid symbols with error bars) and the direct search method (red crosses) from Last Chance Lake [@Axford2017]. Lines show the 1:1 (solid) and 2:1 (dashed) relationships.", fig.width = 80/25.4, fig.height = 80/25.4}
readd(last_chance_plot)
```

The precision of both methods for estimating the count sum from percent data depend on count sum and on the precision to which the percent are recorded (Fig. \@ref(fig:percent-est-plot).
In general, the direct search method is more precise and can give useful estimates when the assemblages have no singletons. 
It can fail if data have been incorrectly rounded, for example by truncation. 
If many digits are given, ignoring the last can solve this.

```{r percent-est-plot, fig.cap = "Accuracy of estimates of the count sum calculated with two different methods with percent data have been rounded to different numbers of digits. Percent data are simulated by resampling a bird data route with 30 species to different count sums.", fig.width = 80/25.4, fig.height = 80/25.4}
readd(percent_rounding_plot)
```

@@

Some of these datasets are older datasets which may, despite the available metadata, have been back-transformed from percent data or digitised with the loss of rare taxa and precision.
For example, all but `r hershop_pollen_summ$has_singleton %>% word()` of the `r hershop_pollen_summ$n` assemblages in the archived pollen data from Hershop Bog, Texas have a greatest common divisor of two or more (the exception has a single singleton and is otherwise divisible by two). 
The original data [@Patty1968] were counted to exactly 200 pollen grains and fern spores. 
It seems likely that the data were converted to percent, rounded and then back-transformed to counts.
For most purposes, such back-transformed or digitised data are perfectly adequate, but the Chao 1 richness estimator calculated using the number of singletons and doubletons, would fail [@Chao1987].

@@

Dataset pollen2 has `r pollen2$n_sample` assemblages, of which `r words(pollen2$n_gcd2)` have a greatest common divisor of two and a further `r words(pollen2$n_gcd3)` with a greatest common divisor of three, one of which has `r pollen2$max_taxa3` taxa.
The probability of having all counts divisible by three in such a taxonomically rich assemblage is 10^`r ceiling(- pollen2$max_taxa3 * log10(3))`.
In addition to the assemblages with a greatest common divisor above one, this dataset includes several assemblages that have a higher proportion of counts divisible by two or three than is typical in assemblages from Neotoma (Fig. \@ref(fig:multiples-plot)).
It is unclear what processes could generate such counts.

```{r multiples-plot, warning = FALSE, fig.height = 100/24.5, fig.width=180/24.5, fig.cap = "A) Distribution of the percent of counts that are divisible by _k_ in each pollen assemblage with at least ten taxa. The vertical dashed line represents the 99.5% percentile. B) Percent of count in each assemblage in dataset pollen2 that are divisible by _k_ against depth. The dashed lines are the quantiles shown in part A. ", message = FALSE}
readd(multiples_plot)
```

In dataset pollen3, the greatest common divisor is three for all assemblages: all `r pollen3$n_count` counts in the dataset are divisible by three (Table \@ref(tab:unexpected-table)). 
For a dataset this large, the probability of having all counts divisible by three by chance is 3^{-`r pollen3$n_count`}^ or 10^`r ceiling(- pollen3$n_count * log10(3))`^.
Given this very small probability, it seems reasonable to conclude that the archived data are not the original counts.

Many assemblages in datasets pollen4 and pollen5 lack singletons and have a greatest common divisor greater than one (Table \@ref(tab:unexpected-table)).
For most of both datasets, which come from the same paper and match the published pollen stratigraphies, assemblages with and without singletons are arranged alternately.
Assemblages without singletons are on the 10 cm intervals, and those with on the intervening 5 cm intervals. 
Closer inspection reveals that the assemblages between two assemblages without singletons have counts that are the mean of their neighbours.
On request, the authors sent the original count data. 
Comparison of these with the archived data showed that the archived assemblages without singletons had had their counts multiplied by two.
The intervening assemblages did not exist in the original data. 
In the published and archived data, they had been interpolated from their neighbours (Fig. \@ref(fig:interpolated-plot)).
The effect of these manipulations is to increase the resolution of the pollen stratigraphy.

```{r unexpected-table}
readd(unexpected_table) %>% knitr::kable(caption = "The number of assemblages, median number of taxa and count sum, percent of assemblages with a greatest common divisor (GCD) of one, in five anonymised pollen datasets.")
```

```{r interpolated-plot, warning = FALSE, fig.height = 100/24.5, fig.width=180/24.5, fig.cap="A) Plot of selected taxa by depth from dataset pollen2. Counts with blue triangles are from assemblages with a greatest common divisor of one. These have been linearly interpolated between the adjacent counts from assemblages with a greatest common divisor of two (red circles). B) Percent of taxa in each assemblage where the count is the mean of the counts in the adjacent assemblages, excluding runs of zero counts. Most of the assemblages with a greatest common divisor of one are exactly linearly interpolated between adjacent assemblages. The main exception is the assemblage at 210 cm which has a single singleton and otherwise has a greatest common divisor of two, other exceptions have a few extra singletons not predicted from the adjacent assembages.", message = FALSE}
readd(pollen4_plot)
```

## Unexpected percent data
```{r line169}

@@

Dataset diatom1 includes `r diatom1_summ$nsamples` diatom assemblages with a median richness of `r diatom1_summ$median_taxa` taxa.
The associated paper reports that at least 400 valves were counted from each assemblage, which means that singletons should have a relative abundance of 0.25% or less.
However, the rarest taxon in two assemblages have a relative abundance of `r diatom1_summ$maxpc`%, and the relative abundance of all other taxa in these assemblages are, within rounding error, integer multiples of this`r diatom1_summ$prop_integer %>% insist(. == 1) %>% invisible()`.
If the count sums actually are at least 400, this would imply that each species count in these assemblages is an integer multiple of at least `r diatom1_summ$integer_mult`.
This seems unlikely.
Alternatively, the actual count sum for these assemblages could be as low as `r diatom1_summ$low_n` valves.
In total, `r diatom1_summ$n_low_n %>% words()` assemblages appear to have a count sum below 400 valves. 
The results section of the paper acknowledges that counts were below the standard 400 valves for part of the Late Glacial.
All the assemblages identified with possible low count sums are in this part of the stratigraphy, demonstrating the utility of the method.

```{r line178}

@@

Dataset diatom2 has `r diatom2_summ$nsamples` assemblages with a median richness of `r diatom2_summ$median_taxa` taxa.
The associated paper reports that the diatom counts included 400–500 valves per assemblage except for four diatom-poor assemblages with at least 100 valves.
The direct search method estimates that `r diatom2_summ$n400 %>% words()` assemblages have a count sum of less than 400, and `r diatom2_summ$n100 %>% words()` have less than 100.
The direct search and minimum percent methods agree, within rounding error, `r diatom2_summ$meth_agree %>% insist(isTRUE(.)) %>% invisible()`on the estimated count sums, implying that either all assemblages have singletons and count sums are low in some assemblages, or that the greatest common divisor of the raw counts is greater than one.
This means that for the assemblage where the minimum percent is `r diatom2_summ$maxpc`, which has `r diatom2_summ$rich_maxpc %>% words()` taxa, all counts would need to be multiples of `r diatom2_summ$integer_mult` for the count sum to be at least 100.

Dataset chironomid1 has `r chironomid1_summ$nsamples` assemblages.
Although the associated paper reports that assemblages with count sums below 50 were discarded, the direct search method estimates that `r chironomid1_summ$n_below50 %>% words` assemblages have count sums between `r chironomid1_summ$min_below50` and `r chironomid1_summ$max_below50`.
These assemblages are diverse (richness between `r chironomid1_summ$richness %>% paste(collapse = " and ")` taxa), so it is unlikely that the true counts are double that estimated here and the greatest common divisor is two.

Palaeoceanographic dataset marine1 has `r marine1_summ$nsamples` assemblages with a median taxonomic richness of `r marine1_summ$median_taxa` taxa.
The associated paper reports that assemblages with count sums below 100 were omitted.
The direct search method estimates that `r marine1_summ$n100 %>% words()` assemblages have count sums below 100 (this excludes `r marine1_summ$n_one_taxa %>% words()` mono-specific assemblage), with count sums as low as `r marine1_summ$est_n_min`.
With this dataset, the direct search and the minimum percent method give identical results for some assemblages but highly divergent estimates for others, with the direct search method sometimes giving estimates of several thousand, in one case over two orders of magnitude higher than the minimum percent method.
Some of the divergence appears to be because some, mainly low diversity, assemblages lack singletons.
The most extreme divergences appear to be data entry errors with incorrect rounding, perhaps during taxonomic revisions of the percent data rather than the raw counts.

@@

```{r}
#set digits to default
options(digits = 7)
```

>>>>

---
title: "Tools for identifying unexpectedly low microfossil count sums"
author:
  - name: Richard J. Telford
    email: richard.telford@uib.no
    affiliation: University of Bergen
    footnote: Corresponding Author
address:
  - code: University of Bergen
    address: Department of Biological Sciences, University of Bergen and Bjerknes Centre for Climate Research, Post Box 7803, N-5020 Bergen, Norway
abstract: |
  Microfossil counts are a key data type in palaeoecology. Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so, and perhaps other aspects of the counts. This paper explores the statistical properties of ecological and palaeoecological species counts and uses the findings to develop methods to flag count or percentage data with unexpected distributions. Several assemblage datasets that fail these tests, and potentially have misreported counts or count sums, are identified.
date: today
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{pdflscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}   
format: pdf
documentclass: article
classoption: a4paper
bibliography: extra/countMS2.bib
csl: extra/elsevier-harvard_rjt.csl
---

```{r}
#| label: setup
#| include: false

# Packages loaded for convenience when knitting without running r_make
library("tidyverse")
library("rjt.misc")

@@

tar_load(bird_singleton_summary, bird_summ)
tar_load(testate_summ)
tar_load(last_chance_summ)
tar_load(pollen_summ, hershop_pollen_summ, wildhorse, pollen_summ1, pollen_summ2, pollen3, pollen1, pollen2)
tar_load(owens_summ1, owens_summ, owens_est_summ)
tar_load(diatom1_summ, diatom2_summ, chironomid1_summ, marine1_summ, chironomid2_summ)

@@

Count data analysed include the 1) North American Breeding Bird survey [@bbsAssistant2], a very large, annual monitoring programme to track the status of North American bird populations.
The survey reports all birds seen in point counts along ~40 km survey routes at the height of the breeding season. 
2) The Brown University planktic foraminiferal database [@Prell1999].
3) The MOLTEN diatom database of modern coastal diatom assemblages [@Clarke2006].
4) The North American testate amoeba training set [@AMESBURY2018].
5) Fossil pollen assemblage from the Neotoma database [@neotoma_2018] downloaded on `r tar_read(pollen_download_date)`.

@@

Across all the datasets, only a small minority of samples lack singletons, and even fewer have a greatest common divisor larger than one (@tbl-multitaxa).
Most of the samples without singletons, or with a greatest common divisor larger than one, have low richness (@fig-richness-singleton-plot).

Although the proportion of pollen assemblages that lack singletons or have a greatest common divisor of one are low, such assemblages are not randomly distributed among the `r pollen_summ$n_datasets |> format()` pollen datasets. 
Half of the assemblages without singletons are in just `r pollen_summ$half_no_sing` datasets.
`r Words(pollen_summ$half_gcd)` datasets contain half of the assemblages with a greatest common divisor above one and only `r pollen_summ$dataset_gcd1`% of the datasets have any such assemblages.
Some pollen datasets with unexpected properties are discussed in section @sec-unexpected-count-data.

```{r}
#| label: tbl-multitaxa
#| tbl-cap: Properties of different data sets.
tar_read(count_summary) |> 
  #manually escape % as cannot use kbl(escape = TRUE) with linebreak
  mutate(across(where(is.character), str_replace, pattern = "%", replacement = "\\\\%")) |> 
  kbl(x = ., booktabs = TRUE, escape = FALSE,
      col.names = linebreak(str_wrap(names(.), width = 10))
    ) |> 
  kable_styling(latex_options = c("basic", "hold_position"))
```

```{r}
#| label: fig-richness-singleton-plot
#| fig.cap: Relationship betwen the presence of singletons and taxonomic richness for
#|   the different datasets. Samples with a greatest common divisor greater than one
#|   are shown in dark blue. Curve is fitted with a generalised additive model
#| fig.height: !expr 120/24.5
#| fig.width: !expr 160/24.5
tar_read(richness_singleton_plot)
```

@@

To explore the effect of taxonomic richness on the prevalence of singletons, I aggregate the bird data to progressively lower taxonomic resolutions.
As richness declines, the proportion of counts without singletons increases (@tbl-multitaxa), reaching a moderate proportion at the order level, where most of the counts lacking singletons have fewer than five taxa.
Even when richness is low, only a small minority of counts have a greatest common divisor greater than one.

@@

To test the sensitivity of the prevalence of counts without singletons to the count sum, I resample the bird counts with 400 or more individuals to smaller count sums using a multivariate hypergeometric distribution.
At the order level, the proportion of counts without singletons decreases steeply until it reaches a minimum at when the count sum is about 200, thereafter it increases slowly (@fig-bird-recount).
The prevalence of counts without singletons is high for small counts because it is possible that only the common taxa are counted; with larger counts, the chance of counting a single individual of a rare taxon increases.
With even larger counts the prevalence of routes without singletons increases due to saturation as singletons become doubletons and few new taxa are found.
The latter effect may largely be an artefact as in the simulation, the list of possible taxa to sample is short, whereas in the wild, there are many rare taxa which might be observed.

```{r}
#| label: fig-bird-recount
#| fig.cap: Effect of count sum on the proportion of bird counts from the North American
#|   breeding bird survey without singleton at the order level. Results are the mean
#|   of ten trials with 95% confidence interval.
#| fig.width: !expr 80/25.4
#| fig.height: !expr 70/25.4
tar_read(bird_recount_singleton_plot)
```

@@

The `r last_chance_summ$nsamples` chironomid head capsule assemblage dataset from Last Chance Lake [@Axford2017] also includes both the count sums and the percentage of each taxa.
The percentages are given to two decimal places: to account for rounding errors, the countSum package uses the smallest and largest values consistent with the reported percentage.
With the minimum percent method the estimated count sums are, within error, either identical to the reported count sum or twice as much (@fig-last-chance-plot).
With the direct search method, the estimated count sums are all exactly twice the reported count sum.
The factor of two difference is because all the counts include half head capsules (but only sometimes is the rarest taxon represented by a half head capsule).
Reporting half microfossils is common for several microfossil groups, including chironomids and pollen (especially for bisaccate conifers); occasionally other fractions are reported.
Half counts make the estimated count too high by a factor of two, so do not risk incorrectly flagging count sums as being too small.

```{r}
#| label: fig-last-chance-plot
#| fig.cap: Estimated and reported chironomid count sums by the minumum percent method
#|   (solid symbols with error bars) and the direct search method (red crosses) from
#|   Last Chance Lake [@Axford2017]. Lines show the 1:1 (solid) and 2:1 (dashed) relationships.
#| fig.width: !expr 80/25.4
#| fig.height: !expr 80/25.4
tar_read(last_chance_plot)
```

The precision of both methods for estimating the count sum from percent data depend on count sum and on the precision to which the percent are recorded (@fig-percent-est-plot).
In general, the direct search method is more precise and can give useful estimates when the assemblages have no singletons. 
It can fail if data have been incorrectly rounded, for example by truncation. 
If many digits are given, ignoring the last can solve this.

```{r}
#| label: fig-percent-est-plot
#| fig.cap: Accuracy of estimates of the count sum calculated with two different methods
#|   with percent data have been rounded to different numbers of digits. Percent data
#|   are simulated by resampling a bird data route with 30 species to different count
#|   sums.
#| fig.width: !expr 80/25.4
#| fig.height: !expr 80/25.4
tar_read(percent_rounding_plot)
```

@@

Some of these datasets are older datasets which may, despite the available metadata, have been back-transformed from percent data or digitised with the loss of rare taxa and precision.
For example, all but `r hershop_pollen_summ$has_singleton |> word()` of the `r hershop_pollen_summ$n` assemblages in the archived pollen data from Hershop Bog, Texas have a greatest common divisor of two or more (the exception has a single singleton and is otherwise divisible by two). 
The original data [@Patty1968] were counted to exactly 200 pollen grains and fern spores. 
It seems likely that the data were converted to percent, rounded and then back-transformed to counts.
For most purposes, such back-transformed or digitised data are perfectly adequate, but the Chao 1 richness estimator calculated using the number of singletons and doubletons, would fail [@Chao1987].

@@

Dataset pollen2 has `r pollen2$n_sample` assemblages, of which `r words(pollen2$n_gcd2)` have a greatest common divisor of two and a further `r words(pollen2$n_gcd3)` with a greatest common divisor of three, one of which has `r pollen2$max_taxa3` taxa.
The probability of having all counts divisible by three in such a taxonomically rich assemblage is 10^`r ceiling(- pollen2$max_taxa3 * log10(3))`.
In addition to the assemblages with a greatest common divisor above one, this dataset includes several assemblages that have a higher proportion of counts divisible by two or three than is typical in assemblages from Neotoma (@fig-multiples-plot).
It is unclear what processes could generate such counts.

```{r}
#| label: fig-multiples-plot
#| warning: false
#| fig.height: !expr 100/24.5
#| fig.width: !expr 180/24.5
#| fig.cap: 'A) Distribution of the percent of counts that are divisible by _k_ in each
#|   pollen assemblage with at least ten taxa. The vertical dashed line represents the
#|   99.5% percentile. B) Percent of count in each assemblage in dataset pollen2 that
#|   are divisible by _k_ against depth. The dashed lines are the quantiles shown in
#|   part A. '
#| message: false
tar_read(multiples_plot)
```

In dataset pollen3, the greatest common divisor is three for all assemblages: all `r pollen3$n_count` counts in the dataset are divisible by three (@tbl-unexpected-table). 
For a dataset this large, the probability of having all counts divisible by three by chance is 3^{-`r pollen3$n_count`}^ or 10^`r ceiling(- pollen3$n_count * log10(3))`^.
Given this very small probability, it seems reasonable to conclude that the archived data are not the original counts.

Many assemblages in datasets pollen4 and pollen5 lack singletons and have a greatest common divisor greater than one (@tbl-unexpected-table).
For most of both datasets, which come from the same paper and match the published pollen stratigraphies, assemblages with and without singletons are arranged alternately.
Assemblages without singletons are on the 10 cm intervals, and those with on the intervening 5 cm intervals. 
Closer inspection reveals that the assemblages between two assemblages without singletons have counts that are the mean of their neighbours.
On request, the authors sent the original count data. 
Comparison of these with the archived data showed that the archived assemblages without singletons had had their counts multiplied by two.
The intervening assemblages did not exist in the original data. 
In the published and archived data, they had been interpolated from their neighbours (@fig-interpolated-plot).
The effect of these manipulations is to increase the resolution of the pollen stratigraphy.

```{r}
#| label: tbl-unexpected-table
#| tbl-cap: The number of assemblages, median number of taxa and count sum, percent of assemblages with a greatest common divisor (GCD) of one, in five anonymised pollen datasets.
tar_read(unexpected_table) |> knitr::kable()
```

```{r}
#| label: fig-interpolated-plot
#| warning: false
#| fig.height: !expr 100/24.5
#| fig.width: !expr 180/24.5
#| fig.cap: A) Plot of selected taxa by depth from dataset pollen2. Counts with blue
#|   triangles are from assemblages with a greatest common divisor of one. These have
#|   been linearly interpolated between the adjacent counts from assemblages with a greatest
#|   common divisor of two (red circles). B) Percent of taxa in each assemblage where
#|   the count is the mean of the counts in the adjacent assemblages, excluding runs
#|   of zero counts. Most of the assemblages with a greatest common divisor of one are
#|   exactly linearly interpolated between adjacent assemblages. The main exception is
#|   the assemblage at 210 cm which has a single singleton and otherwise has a greatest
#|   common divisor of two, other exceptions have a few extra singletons not predicted
#|   from the adjacent assembages.
#| message: false
tar_read(pollen4_plot)
```

## Unexpected percent data
```{r}
#| label: line169

@@

Dataset diatom1 includes `r diatom1_summ$nsamples` diatom assemblages with a median richness of `r diatom1_summ$median_taxa` taxa.
The associated paper reports that at least 400 valves were counted from each assemblage, which means that singletons should have a relative abundance of 0.25% or less.
However, the rarest taxon in two assemblages have a relative abundance of `r diatom1_summ$maxpc`%, and the relative abundance of all other taxa in these assemblages are, within rounding error, integer multiples of this`r diatom1_summ$prop_integer |> insist(. == 1) |> invisible()`.
If the count sums actually are at least 400, this would imply that each species count in these assemblages is an integer multiple of at least `r diatom1_summ$integer_mult`.
This seems unlikely.
Alternatively, the actual count sum for these assemblages could be as low as `r diatom1_summ$low_n` valves.
In total, `r diatom1_summ$n_low_n |> words()` assemblages appear to have a count sum below 400 valves. 
The results section of the paper acknowledges that counts were below the standard 400 valves for part of the Late Glacial.
All the assemblages identified with possible low count sums are in this part of the stratigraphy, demonstrating the utility of the method.

```{r}
#| label: line178

@@

Dataset diatom2 has `r diatom2_summ$nsamples` assemblages with a median richness of `r diatom2_summ$median_taxa` taxa.
The associated paper reports that the diatom counts included 400–500 valves per assemblage except for four diatom-poor assemblages with at least 100 valves.
The direct search method estimates that `r diatom2_summ$n400 |> words()` assemblages have a count sum of less than 400, and `r diatom2_summ$n100 |> words()` have less than 100.
The direct search and minimum percent methods agree, within rounding error, `r diatom2_summ$meth_agree |> insist(isTRUE(.)) |> invisible()`on the estimated count sums, implying that either all assemblages have singletons and count sums are low in some assemblages, or that the greatest common divisor of the raw counts is greater than one.
This means that for the assemblage where the minimum percent is `r diatom2_summ$maxpc`, which has `r diatom2_summ$rich_maxpc |> words()` taxa, all counts would need to be multiples of `r diatom2_summ$integer_mult` for the count sum to be at least 100.

Dataset chironomid1 has `r chironomid1_summ$nsamples` assemblages.
Although the associated paper reports that assemblages with count sums below 50 were discarded, the direct search method estimates that `r chironomid1_summ$n_below50 |> words` assemblages have count sums between `r chironomid1_summ$min_below50` and `r chironomid1_summ$max_below50`.
These assemblages are diverse (richness between `r chironomid1_summ$richness |> paste(collapse = " and ")` taxa), so it is unlikely that the true counts are double that estimated here and the greatest common divisor is two.

Palaeoceanographic dataset marine1 has `r marine1_summ$nsamples` assemblages with a median taxonomic richness of `r marine1_summ$median_taxa` taxa.
The associated paper reports that assemblages with count sums below 100 were omitted.
The direct search method estimates that `r marine1_summ$n100 |> words()` assemblages have count sums below 100 (this excludes `r marine1_summ$n_one_taxa |> words()` mono-specific assemblage), with count sums as low as `r marine1_summ$est_n_min`.
With this dataset, the direct search and the minimum percent method give identical results for some assemblages but highly divergent estimates for others, with the direct search method sometimes giving estimates of several thousand, in one case over two orders of magnitude higher than the minimum percent method.
Some of the divergence appears to be because some, mainly low diversity, assemblages lack singletons.
The most extreme divergences appear to be data entry errors with incorrect rounding, perhaps during taxonomic revisions of the percent data rather than the raw counts.

@@
